msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-14 paper.tex
adds gene conversion section

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{listings}
\usepackage[english]{babel}%
\usepackage[T1]{fontenc}%
\usepackage[utf8]{inputenc}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{dsfont}
\usepackage{verbatim}%
\usepackage{environ}%
\usepackage[right]{lineno}%
%\usepackage{showkeys}

@@

\begin{figure}
\includegraphics[width=\textwidth]{figures/recombination}
\caption{\label{fig-recombination-perf}(A) Approximate run time for \msprime\
simulations as a function of scaled recombination rate.
(B) Comparison of the ARG simulations vs minimal tree sequence representation.}
\end{figure}

In human-like parameter regimes, \msprime's implementation of the
exact coalescent with recombination comprehensively outperforms
all other simulators, including those based on SMC
approximations~\citep{kelleher2016efficient}. However, it is important
to note that although the implementation of Hudson's algorithm is
very efficient, it is still quadratic in the scaled recombination rate
$\rho$. This is because Hudson's algorithms tracks recombinations not only
in segments ancestral to the sample, but also between ancestral segments.
Estimates of the number of such ``trapped'' recombinations grow quadratically
with genome length \citep[Eq.~5.10]{hein2004gene}.
Because they do not track such recombinations, efficient SMC implementations
will be faster that \msprime for sufficiently high recombination rate,
genome length, or $N_e$. %Until we decide to approximate the recombination process
%between distant ancestral segment. Hint hint.

@@

% Seems somewhat unfair to single out Iain and Aylwyn here, but a systematic
% review would be long and tedious, and I'm sure they won't be too upset about
% it.
However, this original definition of the ARG in terms of a stochastic process
is rarely what is meant when the term is currently used. The term has come to
essentially mean a representation of genetic ancestry as a graph; for example,
\cite{mathieson2020ancestry} informally define an ARG as the subset of an
individual's pedigree from which they have inherited ancestral material. Such
imprecision is unfortunate, since there are any number of graphical encodings of
genetic ancestry, and referring to the original Griffiths papers does little to
clarify. As we see below, there are situations in which the precise information
stored in the graph is important, and conflating a high-level
discussion of genetic ancestry with this mathematical object is not
helpful. One may wonder if a more generic term such as ``genetic pedigree''
would be a more useful description, in such cases where the precise form of the
encoding is not important. \jkcomment{Genetic pedigree is probably
just as confusing, on reflection. We can either come up with a different
term, or delete this sentence.}

@@

Gene conversion is \jkcomment{A couple of sentences describing gene
conversion, and what the model is. What other simulators simulate
gc? }
We have implemented this model of gene
conversion in \msprime\ 1.0, and validated the output against
\ms\ and analytical results~\citep{wiuf2000coalescent}.

\begin{figure}
\caption{\label{fig-gc-perf}Comparison of the running time
for E-coli simulations using \msprime\ and SimBac for varying
sample sizes. The parameters used are [] based on [references].}
\end{figure}

Gene conversion is particularly useful to model homologous recombination in
bacterial evolution.
We compare the performance of \msprime\ with gene conversion to
SimBac~\citep{brown2016simbac}, a simulator developed specifically to
simulate bacterial evolution. Although \msprime\ is missing some features
relevant to this case, it is far more efficient at simulating the basic
processes. An important avenue for future work
will be to add further features to model bacterial
evolution. In particular, we aim to include bacterial gene transfer and utilize
the succinct tree sequence structure to represent the resulting ancestral gene
transfer graph~\citep{baumdicker2014AGTG}.

@@

@@

\subsection*{Discrete time Wright-Fisher}
The coalescent is a good approximation for ancient history, but
is bad for the recent past and in particular when we have large
sample
sizes~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}.
Describe basic
approach [DOM, can you fill in please].
See ~\cite{nelson2020accounting} for more details.
We compare the performance of this simulation in Fig X
with ARGON~\citep{palamara2016argon}, a specialised DTWF simulator.

>>>>

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{listings}
\usepackage[english]{babel}%
\usepackage[T1]{fontenc}%
\usepackage[utf8]{inputenc}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{dsfont}
\usepackage{verbatim}%
\usepackage{environ}%
\usepackage[right]{lineno}%
\usepackage{microtype}
%\usepackage{showkeys}

@@

\begin{figure}
% \begin{center}
% \begin{tabular}{cc}
% % \includegraphics[width=5cm]{msprime-scaling.png} &
% % \includegraphics[width=5cm]{figures/recombination} \\
% \end{tabular}
% \end{center}
\includegraphics[width=\textwidth]{figures/recombination}
\caption{\label{fig-recombination-perf}(A) Approximate run time for \msprime\
simulations as a function of scaled recombination rate.
(B) Comparison of the ARG simulations vs minimal tree sequence representation.}
\end{figure}

In human-like parameter regimes, \msprime's implementation of the
exact coalescent with recombination comprehensively outperforms
all other simulators, including those based on SMC
approximations~\citep{kelleher2016efficient}. However, it is important
to note that although the implementation of Hudson's algorithm is
very efficient, it is still quadratic in the scaled recombination rate
$\rho$. This is because Hudson's algorithms tracks recombinations not only
in segments ancestral to the sample, but also between ancestral segments.
Estimates of the number of such ``trapped'' recombinations grow quadratically
with genome length \citep[Eq.~5.10]{hein2004gene}.
Because they do not track such recombinations, efficient SMC implementations
will be faster than \msprime\ for sufficiently high recombination rate,
genome length, or $N_e$. %Until we decide to approximate the recombination process
%between distant ancestral segment. Hint hint.

@@

\jkcomment{If we're going to have this paragraph of ARG bashing then
I guess we'll have to do more than just single out Iain and Aylwyn.}
However, this original definition of the ARG in terms of a stochastic process
is rarely what is meant when the term is currently used. The term has come to
essentially mean any representation of genetic ancestry in the presence
of recombination as a graph-like structure; for example,
\cite{mathieson2020ancestry} informally define an ARG as the subset of an
individual's pedigree from which they have inherited ancestral material. Such
imprecision is unfortunate, since there are any number of graphical encodings of
genetic ancestry, and referring to the original Griffiths papers does little to
clarify. As we see below, there are situations in which the precise information
stored in the graph is important, and conflating a high-level
discussion of genetic ancestry with this mathematical object is not
helpful. The point of technical terms is to provide clarity and
precision on what is being discussed, and it is now essentially
impossible to know what is meant by the term ``ARG'';
we therefore use the term ``genetic ancestry'' in high-level
discussions.

@@

%introduction
Gene conversion is a unidirectional transfer of a short segment of genetic material,
for example between homologous chromosomes~\citep{Chen2007}.
%description of ancestral process
Just as recombination, gene conversion events split the ancestral material
for some lineage, creating two independent lineages.
In contrast to the previously described recombination process, where the ancestral
material to the left and right of a recombination breakpoint follows two
different ancestral lineages, gene conversion events distribute the genetic material
of a short segment to one ancestral lineage and the part to the left and right of this
segment to another lineage.
%Franz Baumdicker: Not sure if this next sentence is necessary,
% but it might be of interest for some people?
The resulting distribution of ancestral material after a gene conversion event
equals the distribution after two recombination events at the left and right position
of the gene conversion tract, followed by a common ancestor event.
%Impact on LD
However, gene conversions impact much shorter segments than crossover recombinations,
typically below 1kb, and thus affects linkage pattern differently than
recombination~\citep{Korunes2017}.

% other simulators
Hudson's \ms\ optionally includes gene conversion and so does
MaCS~\citep{chen2009fast}, while scrm~\citep{staab2014scrm} does not
support gene conversions.
As bacterial recombination can be modeled analogously to eukaryotic allelic
gene conversion, simulators for microbial evolution often include gene conversion
instead of recombination.
This includes the backward simulator SimBac~\citep{brown2016simbac},
the SMC based approach used in FastSimBac~\citep{DeMaio2017},
and forward in time tools within the SLiM framework~\citep{Cury2020}.

In \msprime\ 1.0 we implemented the coalescent with gene conversion as defined by
\cite{wiuf2000coalescent}, where gene conversion events are initiated between any
two sites in the sequence at a constant (scaled) rate $\gamma$ and
the length of the transferred segment is given by a geometric distributed
random variable with mean length $L$.
% Above I used the discrete case, but we could easily switch to a description for the
% continuous case.
We have implemented this model of gene
conversion in \msprime\ 1.0, and validated the output against
\ms\ and analytical results~\citep{wiuf2000coalescent}.

\begin{figure}
\includegraphics[width = \textwidth]{figures/gene_conversion/gc_perf_log.pdf}
% \includegraphics[width = \textwidth]{figures/gene_conversion/gc_perf.pdf}
\caption{\label{fig-gc-perf}Comparison of the running time
for E-coli simulations using \msprime\ and SimBac for varying
sample sizes. The parameters used are chosen to mimic the simulation
of \textit{E. coli} genomes. The genome length was set to 4.5 mb, the scaled per site
gene conversion rate was set to $N_e g = \gamma = 0.015$ and
the mean gene conversion tract length was set to $L = 500$, which is close to
estimated parameters for \textit{E.\ coli}, i.e.\ an effective populations size
$N_e = 1.8e^8$ and a per site per generation conversion rate $g = 8.9e^{-11}$
with mean tract length $L = 542$~\citep{Lapierre2016}.}
\end{figure}

Gene conversion is particularly useful to model homologous recombination in
bacterial evolution.
We compare the performance of \msprime\ with gene conversion to
SimBac~\citep{brown2016simbac}, a simulator developed specifically to
simulate bacterial evolution. Although \msprime\ is missing some features
relevant to this case, e.g.\ the simulation of circular genomes, it is far
more efficient at simulating the basic processes.
Furthermore, gene conversion can be considered in combination
with other current and future features of \msprime\ 1.0,
including varying recombination rates and complex demographies.

An important avenue for future work
will be to add further features to model bacterial
evolution. In particular, we aim to include bacterial gene transfer and utilize
the succinct tree sequence structure to represent the resulting ancestral gene
transfer graph~\citep{baumdicker2014AGTG}.

@@

\begin{figure}
\Large{A FIGURE SHOWING DISCOAL AND MSPRIME}
% \includegraphics[width=\textwidth]{figures/recombination}
\caption{\label{fig-selection-perf}Comparison of the runtime for
\msprime\ and \texttt{discoal} for simulating selective sweeps.
}
\end{figure}

@@

\subsection*{Discrete time Wright-Fisher}
The coalescent is a good approximation for ancient history, but
is bad for the recent past and in particular when we have large
sample
sizes~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}.
Describe basic
approach [DOM, can you fill in please].
See ~\cite{nelson2020accounting} for more details.
We compare the performance of this simulation in Fig.~\ref{fig-dtwf-perf}
with ARGON~\citep{palamara2016argon}, a specialised DTWF simulator.

\begin{figure}
\begin{center}
% \includegraphics[width=6cm]{msprime-argon-runtimes.png}
\end{center}
\caption{\label{fig-dtwf-perf} Comparison of \msprime's DTWF
simulations with ARGON. To ensure the comparison with ARGON is
as fair as possible we run it with a mutation rate of 0 and with
the minimum output options (so that it writes out only an empty VCF file)}
\end{figure}
